---
title: "6260 Final Project Report"
author: "Emily Wu and Cameron Barlow"
date: "2026-04-15"
output: 
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(knitr)
library(ggplot2)
library(MASS)
library(brant)
library(splines)
library(ggeffects)
library(car)
library(corrplot)
```
\begin{titlepage}
    \centering
    \vspace*{5cm}
    {\huge\bfseries 6260 Final Project Report \par}
    \vspace{1.5cm}
    {\Large Emily Wu and Cameron Barlow \par}
    \vfill
    {\large 30 April 2026 \par}
\end{titlepage}

\newpage



# Motivation
Is there a relationship between early life socioeconomic characteristics and family background that can predict the severity of an adult’s criminal history?
We could create the response of criminal history using an ordinal scale to predict the severity part of criminal history using a cumulative logit model.




# Data 
[Talk about sampling]
[Talk about how much data we have available]


The response variable was created through the synthesis of maximum sentence and crime type to model severity
The time variable is highly continuous and skewed in raw form, but much more regular after log transformation, so if ordinal categories are used, they should be created deliberately for interpretability or modeling convenience.



\newpage





```{r, pathway 2 EDA, include=FALSE}
load("37692-0002-Data.rda")
data<-da37692.0002
#1185-1191 for Disorders

v_range <- function(start, end) {
  sprintf("V%04d", as.numeric(start):as.numeric(end))}
MHD<-v_range("1185", "1191")

sblist <- data %>%
  select(all_of(MHD))
sblist<-sblist%>%
  mutate(across(all_of(MHD), ~ as.numeric(gsub(".*\\((-?\\d+)\\).*", "\\1", as.character(.x)))))



sblist <- sblist %>%
  rename("Bipolar" = V1185)%>%
  rename("Depression" = V1186)%>%
  rename("Schizophrenia" = V1187)%>%
  rename("PTSD" = V1188)%>%
  rename("Anxiety" = V1189)%>%
  rename("Personality Disorder" = V1190)%>%
  rename("Other Diagnosis" = V1191)
tables_df <- apply(MARGIN = 2, sblist, FUN = table)
vect<-tables_df[3,]/nrow(data)
tables_df<-rbind(tables_df, vect)

```


```{r, pathway 3 eda}
#proposed variables
varlist<-c("V0062","V0411", "V0413","V0414","V0415","V0935", "V0944", "V0945", "V0982", "V1166", "V1165", "V1168", "V1264", "V1252")
varlist<- c(varlist, v_range("1170", "1174"))

sblist<- data%>%
  dplyr::select(all_of(varlist))



sblist <- sblist %>%
  rename(offense_num = V0062) %>%
  rename("Years of Education" = V0935) %>%
  rename("Total Max Sentence" = V0411) %>%
  rename("Max Sentence # of Years" = V0413) %>%
  rename("Max Sentence # of Months" = V0414) %>%
  rename("Max Sentence # of Days" = V0415) %>%
  rename("SpEd" = V0944) %>%
  rename("Born in the US" = V0945) %>%
  rename("Children" = V0982)%>%
  rename("Foster Home" = V1166)%>%
  rename("Homeless" = V1165)%>%
  rename("Lived with most time" = V1168) %>%
  rename("Welfare" = V1170) %>%
  rename("Public Housing" = V1171) %>%
  rename("Father" = V1172) %>%
  rename("Mother" = V1173) %>%
  rename("Sibling" = V1174) %>%
  rename(SmokeAge = V1252) %>%
  rename(DrinkAge = V1264)
library(corrplot)



sblist <- sblist %>%
  mutate(across(everything(), ~ as.numeric(gsub(".*\\((-?\\d+)\\).*", "\\1", as.character(.x))))) #refer to codebook if needed

sblist$Tot_time <- (sblist$`Max Sentence # of Years`*365 + sblist$`Max Sentence # of Months`*30+  sblist$`Max Sentence # of Days`)/30.44


sblist <- sblist %>%
  mutate(TimeScore = case_when(
    Tot_time < 12 ~ 1,
    Tot_time > 12 & Tot_time < 36 ~ 2,
    Tot_time > 36 & Tot_time < 84 ~ 3,
    Tot_time > 84 & Tot_time < 180 ~ 4,
    Tot_time > 180 & Tot_time < 360 ~ 4,
    Tot_time >= 360 ~ 5
  ))



#scoring - 1: Less than a year


sblist <- sblist %>%
  mutate(edu_group = case_when(
    `Years of Education` < 12 ~ "< HS",
    `Years of Education` == 12 ~ "HS/GED",
    `Years of Education` > 12 & `Years of Education` < 16 ~ "Some college",
    `Years of Education` >= 16 ~ "College+"
  ))
sblist <- sblist %>%
  mutate(offense_type = case_when(
    offense_num == 1 ~ "Violent",
    offense_num == 2 ~ "Property",
    offense_num == 3 ~ "Drug",
    offense_num == 4 ~ "Public Order",
    offense_num == 5 ~ "Other",
    TRUE ~ "Unknown"
  ))
sblist <- sblist %>%
  mutate(CrimeScore = case_when(
    offense_type == "Violent" ~ 5,
    offense_type == "Property" ~ 4,
    offense_type == "Drug" ~ 3,
    offense_type == "Public Order" ~ 2,
    offense_type == "Other" ~ 1
  ))
sblist$TotalScore<- sblist$TimeScore+sblist$CrimeScore



sblist2<-sblist[!is.na(sblist$TotalScore),] #only where data is available
table(sblist2$TotalScore)
colSums(is.na(sblist2))

sblist3 <- sblist2 %>%
  drop_na()
nrow(sblist3)


table(sblist3$TotalScore)





#heavily right skewed

ggplot(sblist, aes(x = log1p(Tot_time))) +
  geom_histogram(fill = "steelblue", color = "white") +
  labs(title = "Distribution of Total Time")

ggplot(sblist, aes(x = Tot_time)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  scale_x_log10() +  
  labs(title = "Log-Scaled Distribution of Total Time (Months)",
       x = "Total time (Log Scale)", 
       y = "Frequency")

ggplot(sblist, aes(x = TotalScore)) +
  geom_histogram(fill = "steelblue", color = "white") +
  labs(title = "Distribution of Total Time")

ggplot(sblist, aes(x = TotalScore)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  scale_x_log10() +  
  labs(title = "Log-Scaled Distribution of Total Time (Months)",
       x = "Total time (Log Scale)", 
       y = "Frequency")
```


# Education Eda/sh
```{r}
#ggplot(sblist2, aes(x = as.factor(`Years of Education`), y = Tot_time, fill = as.factor(`Years of Education`))) +
#  geom_boxplot(outlier.shape = NA, alpha = 0.7) + 
#  theme_minimal() 

heatmap_data <- sblist2 %>%
  filter(!is.na(`Years of Education`), !is.na(TotalScore)) %>%
  group_by(`Years of Education`, TotalScore) %>%
  summarize(count = n(), .groups = "drop")

ggplot(heatmap_data, aes(x = factor(`Years of Education`), y = factor(TotalScore), fill = count)) +
  geom_tile() + 
  scale_fill_viridis_c(option = "magma", direction = -1) + 
  labs(
    title = "Concentration of Inmate Profiles",
    subtitle = "Education vs. TotalScore",
    x = "Years of Education",
    y = "Total Severity Score",
    fill = "# of Inmates"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank()
  )


sblist2 %>%
  count(`Years of Education`, offense_type) %>% # Creates a column named 'n'
  ggplot(aes(x = factor(`Years of Education`), y = factor(offense_type), fill = n)) +
  geom_tile() + 
  scale_fill_viridis_c(option = "magma", direction = -1) + 
  labs(
    title = "Concentration of Inmate Profiles",
    fill = "# of Inmates"
  ) +
  theme_minimal()



ggplot(sblist, aes(x = edu_group, y = log1p(Tot_time))) +
  geom_boxplot() +
  stat_summary(
    fun = median, 
    geom = "text", 
    aes(label = after_stat(round(y, 5))),
    vjust = -0.5,     
    color = "steelblue"
  ) +
  labs(
    x = "Education Group",
    y = "log(1 + Total Time in Prison)",
    title = "Prison Time by Education Group"
  ) +
  theme_minimal()
```



```{r}
sblist2[sblist2$SpEd<0,] #20 observations missing for those who were SpEd or not

ggplot(sblist2, aes(x = SpEd, y = log1p(Tot_time))) +
  geom_jitter(width = 0.2, height = 0.05, alpha = 0.4) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    x = "Special Education",
    y = "log(1 + Total Time in Prison)",
    title = "Special Education Status and Total Time in Prison"
  ) +
  theme_minimal()

```

```{r}
sblist2[sblist2$`Born in the US`<0,] #11 observations missing for those who were born in the US or not

ggplot(sblist2, aes(x = `Born in the US`, y = log1p(Tot_time))) +
  geom_jitter(width = 0.2, height = 0.05, alpha = 0.4) +
  labs(
    x = "Special Education",
    y = "log(1 + Total Time in Prison)",
    title = "Born in the US and Total Time in Prison"
  ) +
  theme_minimal()
```

```{r}
sblist2[sblist2$Children<0,] #19 observations missing for those who were born in the US or not

ggplot(sblist2, aes(x = Children, y = log1p(Tot_time))) +
  geom_jitter(width = 0.2, height = 0.05, alpha = 0.4) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    x = "Special Education",
    y = "log(1 + Total Time in Prison)",
    title = "Born in the US and Total Time in Prison"
  ) +
  theme_minimal()


ggplot(sblist2, aes(x = as.factor(TotalScore), fill = as.factor(SpEd))) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d(option = "mako", begin = 0.2, end = 0.8) + 
  labs(title = "Probability of Rule Violations by Controlling Offense Type",
       x = "Offense Type", 
       y = "Percentage of Inmates",
       fill = "Status") +
  theme_minimal()

sblist2[is.na(sblist2$TotalScore),]
```


```{r, brant testing}
tempy<- data.frame(
  Education = sblist2$`Years of Education`,
  TotalScore = sblist2$TotalScore,
  offense_type = sblist2$offense_type,
  DrinkAge = sblist2$DrinkAge,
  SmokeAge = sblist2$SmokeAge
  
)
model <- polr(as.factor(TotalScore) ~ Education + DrinkAge + SmokeAge, 
              data = tempy, 
              Hess = TRUE)

brant(model)
```


```{r, Multicollinearity Check}
vif_model <- lm(
  as.numeric(TotalScore) ~`Years of Education`+ SpEd+ `Born in the US`+ Children+ `Foster Home`+ Homeless+ Welfare+ `Public Housing`+ Father+ Mother+ Sibling+ SmokeAge+ DrinkAge,
  data = sblist2
)
vif(vif_model)

sub_data <- sblist2 %>% dplyr::select(`Years of Education`, SpEd, `Born in the US`, Children, `Foster Home`, Homeless, Welfare, `Public Housing`, Father, Mother, Sibling, SmokeAge, DrinkAge)
cor_matrix <- cor(sub_data, use = "complete.obs")
corrplot(cor_matrix, method = "color", addCoef.col = "black")

#> vif_model1 <- lm(
#+   as.numeric(TotalScore) ~`Foster Home` + Homeless + Welfare + `Public Housing`,
#+   data = sblist2

#vif_model2 <- lm(
#  as.numeric(TotalScore) ~ Father + Mother + Sibling,
#  data = sblist2)
#vif(vif_model)



```


